Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 numerical-amr.tex
Removed commented-out text from lots of files - basically vestigial junk.
Reformatting tex files so they're all standardized in terms of paragraph wrapping, etc. (no content changed, just formatting.)

%\subsubsection{SAMR Overview}

%Unlike moving mesh methods \citep{1995ApJS..100..269P,1995ApJS...97..231G} or  methods that subdivide individual cells \citep{Adjerid}, Berger \& Collela's AMR (also referred to as \emph{structured} AMR) utilizes an adaptive hierarchy of grid patches at varying levels of resolution.  The SAMR method described here is similar to \citet{Berger89}.  A more complete discussion of the history and development of this type of algorithm is available in \citet{Neeman96}.

@@

In order to simplify bookkeeping, a number of restrictions are imposed on the subgrids:
\begin{itemize}
 \item The refinement factor $r$, or the ratio of the coarse cell width to the fine cell width, must be an integer.  In practice, we generally adopt $r=2$, since tests have shown this typically provides the best performance.
 \item Each subgrid must begin and end on the boundary of a coarse cell.  This means that the number of cells in a patch will be a multiple of the refinement factor.
 \item A subgrid must be completely enclosed by its parent.
\end{itemize}

The AMR grid patches are the primary data structure in \enzo.  Each patch is treated as an individual object that can contain both field variables and particle data.  Individual grids are organized into a  dynamic, distributed hierarchy of mesh patches.  
%See Figure~\ref{fig.2.amrhierdens} for an example of an \enzo\ simulation performed using this AMR scheme where both the baryon density and AMR grid hierarchy are shown together. 
% GB: I think this figure is too application-specific

@@

\begin{figure}
\begin{center}
%\includegraphics[width=0.6\textwidth]{figures/wcycle.eps}
\includegraphics[width=0.6\textwidth]{figures/timestepping.eps}
\end{center}
\caption{\emph{Left:} Example of the timesteps on a 2-level AMR
  hierarchy.  \enzo\ does not restrict the timesteps on the finer levels
  to be a factor of $1/2^n$ smaller.  \emph{Right:} The order in which
  the AMR grids are evaluated on each level.\vspace{1ex}}
\label{fig:wcycle}
\end{figure}

@@

The basic control algorithm looks very much like any used in a grid-based code of a single resolution, and can be written schematically as:
\begin{verbatim}
InitializeHierarchy
While (Time < StopTime):
   dt = ComputeTimeStep(0)
   EvolveLevel(0, dt)
   Time = Time + dt
   CheckForOutput(Time)
\end{verbatim}
The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level upon which to operate as an
argument.  This looks something like the following pseudo-code:
\begin{verbatim}
EvolveLevel(level):
   SetBoundaryValues
   while (Time < ParentTime):
      dt = ComputeTimeStep(level)
      SolveHydroEquations(dt)
      SolveOtherEquations(dt)
      Time = Time + dt
      SetBoundaryValues
      EvolveLevel(level+1, dt)
      FluxCorrection
      Projection
      RebuildHierarchy(level+1)
\end{verbatim}

@@

In the first case, we allow for four types of boundary conditions
(for simplicity, represented here as being at $x=0$, with positive $x$ interior and
negative $x$ exterior to the domain):
\begin{enumerate}
  \item{\em Reflecting.} In this case, the boundary behaves like a mirror, with the solution on the boundary reflecting the solution in the computational domain ($q(-x) = q(+x)$) but with the velocity normal to the boundary reversed: $v_x(-x) = -v_x(+x)$.
  \item{\em Outflow.}  We approximate outflow boundary conditions by duplicating the solution at the edge of the computational domain: $q(-x) = q(0)$.
  \item{\em Inflow.} In this case, the boundary values are fixed by a pre-determined function ($q(-x) = q0(-x,t)$).  The code provides a simple way to set inflow values that are constant in both time and space over an entire face, or hooks are provided for special cases where the boundary must be time-varying.
  \item{\em Periodic.} The boundary solution is obtained from the other side of the grid: $q(-x) = q(x_{\rm{max}}-x)$.
\end{enumerate}

@@

%The same procedure can be used for face-centered quantities; however, for the velocity variable in zeus, we drop the extrema constraint.

@@

Given the grid and its boundary conditions, the process of solving
equations~(\ref{eq:mass})--(\ref{eq:total_energy}) can be detached
from the rest of the SAMR technique, so that it is possible to
experiment with a number of different methods.  The restrictions are
that the method should be flux-conservative, or at least produce an
accurate estimate of the flux of conserved quantities at cell
interfaces.  This is required in the flux-correction step (see
\ref{sec:flux_correction}).  There is also the issue of centering,
which refers to the position of physical quantities within a cell (the
primary types are cell-centered, face-centered, and node-centered). 
%In this description we will employ the direct Eulerian version of the piecewise
%parabolic method \citep{1984JCoPh..54..174C,1995CoPhC..89..149B}, which
%assumes cell-centering for all quantities (actually they are
%cell-averages).  
%However, it is also possible to use other schemes,
%which employ different centering for different variables
%\citep[e.g.,][]{Stone92a}.  
The interpolation step, described in
Section~\ref{sec:interpolation}, must correctly interpolate with the
selected centering.

@@

Since a refined region is simulated with (at least) two different resolutions (one fine and one coarse), it has two or
more ``solutions'' (i.e.~values of density, velocity, etc.) that represent the same volume of space.  This is one important aspect of the SAMR method -- some physical regions are represented by grids at multiple resolutions.  To
maintain consistency, the coarse grid quantities are updated with the finer values, using a volume-weighted average of the conserved quantities:
\begin{equation}
q^{\rm coarse} = r^{-d} \sum q^{\rm fine}_{i, j,k}
\end{equation}
where $r$ is the refinement factor, and the sum is over all fine cells that ``cover" the coarse cell. The exponent, $d$, refers to the weight of the average.  For cell-centered quantities, this is the dimensionality of the problem.  For face-centered quantities, this is one less than the dimensionality, and for edge-centered quantities this is one.

@@

One of the significant advantages of flux-conservative methods is, of course, their ability to maintain conserved quantities to machine precision.  Unfortunately, the projection step upsets this flux conservation since, at the boundary of a refined region, the coarse and fine cells on either side of the border are updated with different estimates of the flux across the boundary.  At one refinement level this comes from the coarse grid computation, and in the other from the fine grid solution.  Conservation can be restored by correcting all coarse cells that lie outside the boundary of a fine region
with the difference between the coarse and fine estimates of the flux
across the boundary:
\begin{equation}
  q^{\rm coarse} = \tilde{q}^{\rm coarse} - \Delta t \left( F^{\rm
      coarse} - \sum_{j,k} F^{\rm fine}_{j,k} \right)
\end{equation}
where $F$ is the flux of quantity $q_i$ in the $i$-direction through the cell interface.   We use $\tilde{q}$ to indicate the uncorrected quantity on the coarse grid and the sum is over the $r^{d-1}$ fine cell interfaces in the perpendicular dimensions that share a face with the coarse cell being corrected.  This correction is carried out for all coarse cells that share a face with a refined cell (and are not themselves covered with more refined cells).

@@

\item{\em Baryon mass.}  This refinement criterion is designed to
  mimic a Lagrangian method in that it tries to keep a fixed mass
  resolution.  In this method, a cell is refined if the baryonic mass
  in the cell ($M_g = \rho\, (\Delta x)^d$) is larger than a specified value:
\begin{equation}
M_g > \rho_{\rm flag}\, (\Delta x_{\rm root})^d r^{\epsilon_l l},
\end{equation}
where $\rho_{\rm flag}$ is the equivalent density on the root grid
required for refinement, $\Delta x_{\rm root}$ is the root grid cell
spacing, $r$ is the refinement factor, $l$ is the level and
$\epsilon_l$ is a parameter that can be used to make the refinement
more aggressive ($\epsilon_l < 0$) or less aggressive ($\epsilon_l >
0$) than pure Lagrangian refinement --- in other words,
super-Lagrangian or sub-Lagrangian refinement.

\item{\em Shocks.}  In this method, we identify and refine cells that contain shocks using the following criteria:
\begin{eqnarray}
%% (p_{i+1} - p_{i-1})/\min(p_{i+1}, p_{i-1}) & > & 0.33,  \nonumber \\
%% u_{i-1} - u_{i+1} & > & 0, \quad\text{and} \\
%% e_i / E_i & > & 0.1.  \nonumber
(p_{i+1} - p_{i-1})/\min(p_{i+1}, p_{i-1}) > 0.33,  \qquad
u_{i-1} - u_{i+1} > 0, \quad\text{and} \qquad
e_i / E_i > 0.1.  \nonumber
\end{eqnarray}
We note that the parameters $0.33$ and $0.1$ in the equations above were determined empirically, and can be modified by the user to refine more or less aggressively, as the situation warrants.

@@

\item{\em Jeans length.}  Following \cite{Truelove98}, the code can be forced to refine the Jeans length by a fixed number of cells, or, more specifically, whenever the following criterion is not met:
\begin{equation}
\Delta x < \left( \frac{\pi k_B T}{N_J^2 G \rho \mh} \right)^{1/2},
\end{equation}
where $J_J$ is the required number of cells per Jeans length (4 by default).

@@

\item{\em Shear.} As described in \citet{Kritsuk06}, turbulence calculations can be successfully modeled with AMR if the shear is used as a refinement parameter; in particular, refinement occurs when the finite difference version of the following inequality holds:
\begin{equation}
%\left(\frac{du}{dy}\right)^2 + \left(\frac{du}{dz}\right)^2 + 
%\left(\frac{dv}{dx}\right)^2 + \left(\frac{dv}{dz}\right)^2 + 
%\left(\frac{dw}{dx}\right)^2 + \left(\frac{dw}{dy}\right)^2 
\sum_i \sum_j \left( \frac{\partial v_i}{\partial x_j} \right)^2 -  \sum_i \left( \frac{\partial v_i}{\partial x_i} \right)^2
> \epsilon_s \frac{c_s^2}{\Delta x^2},
\end{equation}
where $\epsilon_s$ is a dimensionless parameter.

\item{\em Optical depth.} This refinement criterion, typically used for simulations using radiation transport, 
flags cells for refinement if they have an \ion{H}{1} optical depth lower than unity.  The optical depth of a cell is estimated to be:
\begin{equation}
\tau = \sigma_{\rm HI} n_{\rm HI} \Delta x 
\end{equation}
where $\sigma_{\rm HI}$ is the neutral atomic hydrogen absorption
cross section, $n_{\rm HI}$ is the proper \ion{H}{1} number density,
and $\Delta x$ is the cell width (which is assumed to be the same along each axis).  If $\tau > 1$, the cell is refined
further, up to the maximum level of refinement.

\item{\em Resistive length.}  In a way similar to the Jeans length criterion described above, the code can be forced to locally resolve the resistive length scale by a minimum of a fixed number of cells.  This is done by ensuring that the resistive length scale, defined as
\begin{equation}
L_R = |B| / |\curl B|
\end{equation}
is always refined by at least N$_R$ cells locally (i.e.~we refine a given cell further if $L_R \leq N_R \Delta x$).  Note that \enzo\ does not actually solve the resistive MHD equations -- this is the ``characteristic'' local resistive length, rather than one calculated within the MHD solver.

@@

%GB: The following section has now been worked into the main text above and, I think, can be removed.
%
%\subsection{Adaptive mesh refinement}\label{sec.ov.amr}
%\dcc{removed most of the notes.  Added intro blurb.  moved parallelism
%  stuff. Re-worded the bit on ghost zones.  Summary of 'grid as
%  independent update + syncing with neighbors and parents,' W-cycle. Added
%  sentence on subgrid creation.  Made timestepping its own subsection.}
%\red{Talk about reasonable refinement factors - which 
%ones are usable, etc?}

%The primary purpose of the \enzo\ code is its Adaptive Mesh Refinement
%capability, which allows it to reach extremely large dynamical ranges
%with limited computational resources, opening doors previously closed
%by finite memory and computational time. Unlike moving mesh methods
%\citep{1995ApJS..100..269P,1995ApJS...97..231G} or  
%methods that subdivide 
%individual cells \citep{Adjerid}, Berger \& Collela's AMR (also referred 
%to as \emph{structured} AMR) utilizes an adaptive hierarchy of grid 
%patches at varying levels of resolution.  Each rectangular grid patch 
%(referred to as a ``grid'') covers some region of space in its 
%\emph{parent grid} which requires higher resolution, and can itself 
%become the parent grid to an even more highly resolved \emph{child grid}. 
%ENZO's implementation of structured AMR places no restrictions on 
%the number of grids at a given level of refinement, or on the number of 
%levels of refinement.  However, owing to limited computational resources 
%it is practical to institute a maximum level of refinement $\Lmax$.

%The \enzo\ implementation of AMR allows arbitrary integer ratios of parent and
%child grid resolution.  However, in practice 
%refinement factors (defined as the ratio of parent grid resolution to child grid resolution) 
%of two or four are typically used, with a refinement factor
%of two being most commonly used for cosmological simulations for reasons of
%efficiency. The ratio of
%box size to the maximum grid resolution of a given simulation is
%therefore $L/e = \Nroot \times 2^{\Lmax}$, where $\Nroot$ is the
%number of cells along one edge of the root grid, $\Lmax$ is the
%maximum level of refinement allowed, and L and e are the box size
%and grid resolution of the most highly refined region, respectively.

%The AMR grid patches are the primary data structure in \enzo.  Each
%patch is treated as an individual object which can contain both field
%variables and particle data.  Individual grids are organized into a 
%dynamic, distributed hierarchy of mesh patches.  
%See 
%Figure~\ref{fig.2.amrhierdens} for an example of an \enzo\
%simulation performed using this AMR scheme where both the baryon density
%and AMR grid hierarchy are shown together.

%\begin{figure}
%\begin{center}
%\includegraphics[width=0.4\textwidth]{figures/amr-hier-dens.eps}
%\caption{Example of an \enzo\ simulation showing the AMR grid hierarchy.  
%This is a simulation of the collapse of a single massive halo with a 
%$32^3$ root grid and two $32^3$ static nested grids.  AMR is only allowed
%within the most highly refined static nested grid.  Log baryon density is shown
%by colored isocontours with values corresponding to the legend at center left.  
%The rectangular solid wire frames correspond to individual 
%AMR grids, with different colors corresponding to level as indicated by the
%legend at top left.}
%\label{fig.2.amrhierdens}
%\end{center}
%\end{figure}

%Each grid patch in \enzo\ contains arrays of values for baryon and 
%particle quantities.   Grids are 
%partitioned into a core of \emph{real zones} and a surrounding layer 
%of \emph{ghost zones}.  The real zones store field values and ghost 
%zones are used to temporarily store values which have been obtained 
%directly from neighboring grids or interpolated from a parent grid.  
%These zones are necessary to accommodate the computational stencil of 
%the hydrodynamics solvers (Sections~\ref{sec.ov.hydro.ppm} and~\ref{sec.ov.hydro.zeus}) 
%and the gravity solver (Section~\ref{sec.ov.nbody}).  
%The PPM and \zeus\ hydro solvers require 3 layers of ghost zones,  and the gravity
%solver requires 6.  This can lead to significant memory 
%and computational overhead, particularly for smaller grid patches
%at high levels of refinement.  More information on filling of ghost
%zones can be found in the appendix.

%From the point of view of the hydro solver, each grid is solved as an
%independent CFD problem, with Dirichlet boundary conditions stored in
%the ghost zones.  At the end of each timestep, cells on a given level
%fill ghost zones by interpolating from the parent and copying from
%neighbors on the same level.  Grids then correct the flux at the interface boundary between
%fine and coarse zones, and finally project their real zone data to the
%parent grid.

%Timesteps are determined on each level as
%described in \ref{sed.timestepping}, and the hierarchy is advanced on
%a level by level basis in what's known as a W-Cycle.    Beginning with
%the coarsest level, $L$, all grids on that level are advanced one
%timestep.  Then one timestep is taken on all grids at the next level
%of refinement, $L+1$, and so on until the finest level is resolved.
%The finest level is then advanced, using as many steps as it takes to
%reach the level immediately above.  The finest level is then synced
%to the level above it, which then proceeds forewards one more step.
%\red{Needs a picture of the W cycle.}

%At the end of every timestep on every level each grid updates 
%its ghost zones by exchanging information with its neighboring grid patches 
%(if any exist) and/or by interpolating from a parent grid.  
%In addition, cells are examined to see if they should be refined or 
%de-refined, and the entire grid hierarchy is rebuilt at that 
%level (including all more highly refined levels).  The timestepping and 
%hierarchy advancement/rebuilding process described here is repeated 
%recursively on every level to the specified maximum level of 
%refinement in the simulation.

%Since the addition of more highly refined grids is adaptive,  
%the conditions for refinement must be specified.  The criteria of 
%refinement can be set by the threshold value of the overdensity of 
%baryon gas or dark matter in a cell (which is really a refinement on
%the mass of gas or DM in a cell), the local Jeans length, the 
%local density gradient, or local pressure and energy gradients.  
%A cell reaching any or all of these criteria 
%will then be flagged for refinement.  
%Once all cells at a given level have been examined, rectangular boundaries 
%are determined which minimally encompass the flagged cells. The set of
%flagging cells is then further fractured into smaller bounding
%rectangles to ensure computational 
%efficiency.  More details on subgrid creation can be found in Appendix
%\ref{app.subgrid_creation}.  A refined grid patch is 
%introduced within each such bounding rectangle. Thus the cells needing 
%refinement, as well as adjacent cells within the patch which do not need 
%refinement, are refined. While this approach is not as memory efficient as 
%cell-splitting AMR schemes, it offers more freedom with 
%finite difference stencils. For example, PPM requires a stencil of seven 
%cells per dimension. This cannot easily be accommodated in cell-splitting 
%AMR schemes.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ms"
%%% End:

>>>>

@@

In order to simplify bookkeeping, a number of restrictions are imposed
on the subgrids:

\begin{itemize}
 \item The refinement factor $r$, or the ratio of the coarse cell
   width to the fine cell width, must be an integer.  In practice, we
   generally adopt $r=2$, since tests have shown this typically
   provides the best performance.
 \item Each subgrid must begin and end on the boundary of a coarse
   cell.  This means that the number of cells in a patch will be a
   multiple of the refinement factor.
 \item A subgrid must be completely enclosed by its parent.
\end{itemize}

The AMR grid patches are the primary data structure in \enzo.  Each
patch is treated as an individual object that can contain both field
variables and particle data.  Individual grids are organized into a
dynamic, distributed hierarchy of mesh patches.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/timestepping.eps}
\end{center}
\caption{\emph{Left:} Example of the timesteps on a 2-level AMR
  hierarchy.  \enzo\ does not restrict the timesteps on the finer levels
  to be a factor of $1/2^n$ smaller.  \emph{Right:} The order in which
  the AMR grids are evaluated on each level.\vspace{1ex}}
\label{fig:wcycle}
\end{figure}

@@

The basic control algorithm looks very much like any used in a
grid-based code of a single resolution, and can be written
schematically as:

\begin{verbatim}
InitializeHierarchy
While (Time < StopTime):
   dt = ComputeTimeStep(0)
   EvolveLevel(0, dt)
   Time = Time + dt
   CheckForOutput(Time)
\end{verbatim}

The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level upon which to operate as
an argument.  This looks something like the following pseudo-code:

\begin{verbatim}
EvolveLevel(level):
   SetBoundaryValues
   while (Time < ParentTime):
      dt = ComputeTimeStep(level)
      SolveHydroEquations(dt)
      SolveOtherEquations(dt)
      Time = Time + dt
      SetBoundaryValues
      EvolveLevel(level+1, dt)
      FluxCorrection
      Projection
      RebuildHierarchy(level+1)
\end{verbatim}

@@

In the first case, we allow for four types of boundary conditions (for
simplicity, represented here as being at $x=0$, with positive $x$
interior and negative $x$ exterior to the domain):

\begin{enumerate}
  \item{\em Reflecting.} In this case, the boundary behaves like a
    mirror, with the solution on the boundary reflecting the solution
    in the computational domain ($q(-x) = q(+x)$) but with the
    velocity normal to the boundary reversed: $v_x(-x) = -v_x(+x)$.
  \item{\em Outflow.}  We approximate outflow boundary conditions by
    duplicating the solution at the edge of the computational domain:
    $q(-x) = q(0)$.
  \item{\em Inflow.} In this case, the boundary values are fixed by a
    pre-determined function ($q(-x) = q0(-x,t)$).  The code provides a
    simple way to set inflow values that are constant in both time and
    space over an entire face, or hooks are provided for special cases
    where the boundary must be time-varying.
  \item{\em Periodic.} The boundary solution is obtained from the
    other side of the grid: $q(-x) = q(x_{\rm{max}}-x)$.
\end{enumerate}

@@

@@

Given the grid and its boundary conditions, the process of solving
equations~(\ref{eq:mass})--(\ref{eq:total_energy}) can be detached
from the rest of the SAMR technique, so that it is possible to
experiment with a number of different methods.  The restrictions are
that the method should be flux-conservative, or at least produce an
accurate estimate of the flux of conserved quantities at cell
interfaces.  This is required in the flux-correction step (see
\ref{sec:flux_correction}).  There is also the issue of centering,
which refers to the position of physical quantities within a cell (the
primary types are cell-centered, face-centered, and node-centered).
The interpolation step, described in Section~\ref{sec:interpolation},
must correctly interpolate with the selected centering.

@@

Since a refined region is simulated with (at least) two different
resolutions (one fine and one coarse), it has two or more
``solutions'' (i.e.~values of density, velocity, etc.) that represent
the same volume of space.  This is one important aspect of the SAMR
method -- some physical regions are represented by grids at multiple
resolutions.  To maintain consistency, the coarse grid quantities are
updated with the finer values, using a volume-weighted average of the
conserved quantities:

\begin{equation}
q^{\rm coarse} = r^{-d} \sum q^{\rm fine}_{i, j,k}
\end{equation}

where $r$ is the refinement factor, and the sum is over all fine cells
that ``cover" the coarse cell. The exponent, $d$, refers to the weight
of the average.  For cell-centered quantities, this is the
dimensionality of the problem.  For face-centered quantities, this is
one less than the dimensionality, and for edge-centered quantities
this is one.

@@

One of the significant advantages of flux-conservative methods is, of
course, their ability to maintain conserved quantities to machine
precision.  Unfortunately, the projection step upsets this flux
conservation since, at the boundary of a refined region, the coarse
and fine cells on either side of the border are updated with different
estimates of the flux across the boundary.  At one refinement level
this comes from the coarse grid computation, and in the other from the
fine grid solution.  Conservation can be restored by correcting all
coarse cells that lie outside the boundary of a fine region with the
difference between the coarse and fine estimates of the flux across
the boundary:

\begin{equation}
  q^{\rm coarse} = \tilde{q}^{\rm coarse} - \Delta t \left( F^{\rm
      coarse} - \sum_{j,k} F^{\rm fine}_{j,k} \right)
\end{equation}

where $F$ is the flux of quantity $q_i$ in the $i$-direction through
the cell interface.   We use $\tilde{q}$ to indicate the uncorrected
quantity on the coarse grid and the sum is over the $r^{d-1}$ fine
cell interfaces in the perpendicular dimensions that share a face with
the coarse cell being corrected.  This correction is carried out for
all coarse cells that share a face with a refined cell (and are not
themselves covered with more refined cells).

@@

\item{\em Baryon mass.}  This refinement criterion is designed to
  mimic a Lagrangian method in that it tries to keep a fixed mass
  resolution.  In this method, a cell is refined if the baryonic mass
  in the cell ($M_g = \rho\, (\Delta x)^d$) is larger than a specified value:

\begin{equation}
M_g > \rho_{\rm flag}\, (\Delta x_{\rm root})^d r^{\epsilon_l l},
\end{equation}

where $\rho_{\rm flag}$ is the equivalent density on the root grid
required for refinement, $\Delta x_{\rm root}$ is the root grid cell
spacing, $r$ is the refinement factor, $l$ is the level and
$\epsilon_l$ is a parameter that can be used to make the refinement
more aggressive ($\epsilon_l < 0$) or less aggressive ($\epsilon_l >
0$) than pure Lagrangian refinement --- in other words,
super-Lagrangian or sub-Lagrangian refinement.

\item{\em Shocks.}  In this method, we identify and refine cells that
  contain shocks using the following criteria:

\begin{eqnarray}
(p_{i+1} - p_{i-1})/\min(p_{i+1}, p_{i-1}) > 0.33,  \qquad
u_{i-1} - u_{i+1} > 0, \quad\text{and} \qquad
e_i / E_i > 0.1.  \nonumber
\end{eqnarray}

We note that the parameters $0.33$ and $0.1$ in the equations above
were determined empirically, and can be modified by the user to refine
more or less aggressively, as the situation warrants.

@@

\item{\em Jeans length.}  Following \cite{Truelove98}, the code can be
  forced to refine the Jeans length by a fixed number of cells, or,
  more specifically, whenever the following criterion is not met:

\begin{equation}
\Delta x < \left( \frac{\pi k_B T}{N_J^2 G \rho \mh} \right)^{1/2},
\end{equation}
where $J_J$ is the required number of cells per Jeans length (4 by default).

@@

\item{\em Shear.} As described in \citet{Kritsuk06}, turbulence
  calculations can be successfully modeled with AMR if the shear is
  used as a refinement parameter; in particular, refinement occurs
  when the finite difference version of the following inequality
  holds:

\begin{equation}
\sum_i \sum_j \left( \frac{\partial v_i}{\partial x_j} \right)^2 -  \sum_i \left( \frac{\partial v_i}{\partial x_i} \right)^2
> \epsilon_s \frac{c_s^2}{\Delta x^2},
\end{equation}
where $\epsilon_s$ is a dimensionless parameter.

\item{\em Optical depth.} This refinement criterion, typically used
for simulations using radiation transport, flags cells for refinement
if they have an \ion{H}{1} optical depth lower than unity.  The
optical depth of a cell is estimated to be:

\begin{equation}
\tau = \sigma_{\rm HI} n_{\rm HI} \Delta x 
\end{equation}

where $\sigma_{\rm HI}$ is the neutral atomic hydrogen absorption
cross section, $n_{\rm HI}$ is the proper \ion{H}{1} number density,
and $\Delta x$ is the cell width (which is assumed to be the same
along each axis).  If $\tau > 1$, the cell is refined further, up to
the maximum level of refinement.

\item{\em Resistive length.}  In a way similar to the Jeans length
  criterion described above, the code can be forced to locally resolve
  the resistive length scale by a minimum of a fixed number of cells.
  This is done by ensuring that the resistive length scale, defined as

\begin{equation}
L_R = |B| / |\curl B|
\end{equation}

is always refined by at least N$_R$ cells locally (i.e.~we refine a
given cell further if $L_R \leq N_R \Delta x$).  Note that \enzo\ does
not actually solve the resistive MHD equations -- this is the
``characteristic'' local resistive length, rather than one calculated
within the MHD solver.

@@

### 2013-06-25 numerical-chemistry.tex
Removed commented-out text from lots of files - basically vestigial junk.
Reformatting tex files so they're all standardized in terms of paragraph wrapping, etc. (no content changed, just formatting.)

\enzo\ solves the rate equations using a method based on a semi-implicit
formulation in order to provide a stable, positive definite and first-order
accurate solution.  The update discretization splits chemical changes into
formation components and destruction components and updates with a mixed set of
time states, as described in \citet{anninos97}.  The formation components of
species $S_i$ are computed at the current subcycle time, where the contribution
of species $S_i$ to its own destruction components are computed at the updated
time; all other contributions to the destruction component are computed at the
current time.  This mixed state improves accuracy, ensures species values
are positive definite, and is equivalent to one Jacobi iteration of an implicit
Euler solve.  This technique is optimized by taking the chemical intermediaries
\Hm and \HHp, which have large rate coefficients and low concentrations, and
grouping them into a separate category of equations.  Due to their fast
reactions, these species are very sensitive to small changes in the more
abundant species and are (at almost all times in astrophysical calculations) close to
their equilibrium values.  Attempting to resolve their formation and destruction
times would necessitate extremely small timesteps.
%and while these species
%provide necessary channels to facilitate the formation of molecular hydrogen at
%low densities, they do not otherwise significantly influence the abundance of
%species with higher concentrations.  
Therefore, reactions governing these two
species can be decoupled from the rest of the network and treated independently
through analytic solutions for equilibrium values.  This allows a significant speedup
in the solution speed, as the timestepping scheme is applied only to the slower 7- or
10-species network (depending on whether deuterium is included or not), which
will be much closer to the overall hydrodynamic timestep of the simulation.

>>>>

\enzo\ solves the rate equations using a method based on a
semi-implicit formulation in order to provide a stable, positive
definite and first-order accurate solution.  The update discretization
splits chemical changes into formation components and destruction
components and updates with a mixed set of time states, as described
in \citet{anninos97}.  The formation components of species $S_i$ are
computed at the current subcycle time, where the contribution of
species $S_i$ to its own destruction components are computed at the
updated time; all other contributions to the destruction component are
computed at the current time.  This mixed state improves accuracy,
ensures species values are positive definite, and is equivalent to one
Jacobi iteration of an implicit Euler solve.  This technique is
optimized by taking the chemical intermediaries \Hm and \HHp, which
have large rate coefficients and low concentrations, and grouping them
into a separate category of equations.  Due to their fast reactions,
these species are very sensitive to small changes in the more abundant
species and are (at almost all times in astrophysical calculations)
close to their equilibrium values.  Attempting to resolve their
formation and destruction times would necessitate extremely small
timesteps.  Therefore, reactions governing these two species can be
decoupled from the rest of the network and treated independently
through analytic solutions for equilibrium values.  This allows a
significant speedup in the solution speed, as the timestepping scheme
is applied only to the slower 7- or 10-species network (depending on
whether deuterium is included or not), which will be much closer to
the overall hydrodynamic timestep of the simulation.

### 2013-06-25 numerical-conduction.tex
Reformatting tex files so they're all standardized in terms of paragraph wrapping, etc. (no content changed, just formatting.)

\enzo\ implements the equations of isotropic heat conduction in a manner
similar to that of \citet{2007ApJ...664..135P}.  
The isotropic flux of heat is given by equation~(\ref{eq:conduction}) and we
use a value for the Spitzer conduction coefficient, 
$\kappa_{\rm sp} = 4.6 \times 10^{-7}$~T$^{5/2}$ erg
s$^{-1}$~cm$^{-1}$~K$^{-1}$ \citep{1962pfig.book.....S}.  In this
situation we are using a
value for the Coulomb logarithm, $\log \Lambda = 37.8$, that is
appropriate for the intracluster medium \citep{1988xrec.book.....S} --
in astrophysically-relevant, fully ionized plasmas this value varies
by no more than 50\% (see, e.g., Smith et al. 2013, submitted).
It is quite possible that the local heat flux computed in this way 
can become unphysically large in
the high-temperature, low-density cluster regime when using this
formulation; therefore, we take into account the saturation of the heat
flux \citep{1977ApJ...211..135C} at a maximum level of
\begin{equation}
F_{sat} \simeq 0.4 n_e k_b T \left( \frac{2 k_b T}{\pi m_e} \right)^{1/2}.
\end{equation}
To ensure a smooth transition between
the Spitzer and saturated regimes, we define an effective
conductivity using the formalism of \citet{1988xrec.book.....S}
\begin{equation}
\kappa_{eff} = \frac{\kappa_{\rm cond}}{1 + 4.2 \lambda_e / \ell_T} \; ,
\end{equation}
where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\grad T|$ is the characteristic length scale of the local
temperature gradient.  We also assume that the conductivity of the
plasma can be described in terms of an effective conductivity, which
can be expressed as a fraction f$_{\rm sp}$ of the Spitzer conductivity
(where f$_{\rm sp} \leq 1.0$ are considered physically realistic values).  This takes
into account physical processes below the resolution limit of the
simulation, such as tangled magnetic fields, that can suppress heat
transport.

>>>>

\enzo\ implements the equations of isotropic heat conduction in a
manner similar to that of \citet{2007ApJ...664..135P}.  The isotropic
flux of heat is given by equation~(\ref{eq:conduction}) and we use a
value for the Spitzer conduction coefficient, $\kappa_{\rm sp} = 4.6
\times 10^{-7}$~T$^{5/2}$ erg s$^{-1}$~cm$^{-1}$~K$^{-1}$
\citep{1962pfig.book.....S}.  In this situation we are using a value
for the Coulomb logarithm, $\log \Lambda = 37.8$, that is appropriate
for the intracluster medium \citep{1988xrec.book.....S} -- in
astrophysically-relevant, fully ionized plasmas this value varies by
no more than 50\% (see, e.g., Smith et al. 2013, submitted).  It is
quite possible that the local heat flux computed in this way can
become unphysically large in the high-temperature, low-density cluster
regime when using this formulation; therefore, we take into account
the saturation of the heat flux \citep{1977ApJ...211..135C} at a
maximum level of

\begin{equation}
F_{sat} \simeq 0.4 n_e k_b T \left( \frac{2 k_b T}{\pi m_e} \right)^{1/2}.
\end{equation}

To ensure a smooth transition between the Spitzer and saturated
regimes, we define an effective conductivity using the formalism of
\citet{1988xrec.book.....S}

\begin{equation}
\kappa_{eff} = \frac{\kappa_{\rm cond}}{1 + 4.2 \lambda_e / \ell_T} \; ,
\end{equation}

where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\grad T|$ is the characteristic length scale of the local
temperature gradient.  We also assume that the conductivity of the
plasma can be described in terms of an effective conductivity, which
can be expressed as a fraction f$_{\rm sp}$ of the Spitzer
conductivity (where f$_{\rm sp} \leq 1.0$ are considered physically
realistic values).  This takes into account physical processes below
the resolution limit of the simulation, such as tangled magnetic
fields, that can suppress heat transport.
